The state of knowledge for marine invertebrate biodiversity in the continental US

### 2018-04-16 manuscript.Rmd
fix small typos before incorporating GP's edits

---
#author: Fran\c{c}ois Michonneau
title: "The state of knowledge for marine invertebrate biodiversity in the continental US"
date: ' ' #'`r format(Sys.time())`'
output:
  bookdown::pdf_document2:
    includes:
    #  before_body: ../template/doc_prefix.tex
      in_header: preamble.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc: no
fontsize: 12pt
link-citations: yes
csl: nature.csl
bibliography: /home/francois/Library/dark-plankton.bib
abstract: Museum collections have been recording our knowledge about species and their geographic ranges for about 250 years, but until recently, this information was fragmented across institutions and difficult to synthesize. The digitization of biodiversity information is now enabling the synthesis of this information, and can be used to highlight knowledge gaps. Here we use publicly accessible biodiversity data to assemble a list of `r format(n_distinct(combined_species$worms_valid_name), big.mark = ",")` species of marine invertebrates from the continental United States. Our knowledge about species richness and their geographic ranges is limited by sampling effort, as `r prop_10_specimens_species`% of the species are known from less than 10 occurrences. While recent developments in high-throughput sequencing promise to greatly speed up the documentation of marine biodiversity, anchoring results from such studies into existing DNA sequence data and associated vouchers would provide a richer biological context to interpret the insights provided by these datasets. However, with `r format_output(prop_spp_barcoded$p_barcoded * 100)`% of the marine invertebrate fauna that has sequences available for the most widely used DNA barcoding gene (cytochrome oxidase I, COI), and given that for many species these records have been sampled in biogeographic areas outside the USA, additional efforts are needed to assemble a comprehensive library of reference DNA barcoding sequences. More sampling, coordinated at the national scale, is needed to improve our knowledge of species and their geographic ranges. This data will allow the detection of the impacts of climate change and other anthropogenic strains on marine biodiversity. Advances in computing, information systems, and sequencing technologies, have set the stage to inform where future efforts for documenting biodiversity should be applied.
---

@@

Several factors may contribute to species being included in the regional lists but not having records in biodiversity databases. The first category of biases is linked to the likelihood having specimens for these species in digitized museum collections. Rare species may only have been collected a few times, and the corresponding records have not been digitized yet. Species that are difficult to preserve because they are large, very small, or fragile, are deposited in museums less frequently. The other category of biases is associated with data quality for these records. Some of these species may be in the databases but do not have geographic coordinates associated with them, and therefore would not be included in the results of our query. While it is difficult to assess biases associated with the likelihood of preservation of a species, we investigated how missing geographic coordinates are affecting the discrepancies between the checklists and the databases.

@@

The lack of tools for the authors of these checklists to have access to museum records across institutions explains why some species were not included in the checklists. Many of these records were difficult to discover before digitization which, for the most part, has occurred after the compilation of these checklists. We estimate that using biodiversity databases increases the number of species included in the checklist by `r format_output(spp_not_in_lists$gom_not_in_list/n_distinct(total_gom_species$worms_valid_name)*100)`% for the Gulf of Mexico, and `r format_output(spp_not_in_lists$pnw_not_in_list/n_distinct(total_pnw_species$worms_valid_name)*100)`% for the Pacific Northwest.

@@

The marine invertebrates of the US have generally been characterized genetically at a higher level than other parts of the world: `r format_output(prop_spp_barcoded$p_barcoded * 100)`% (n = `r format(nrow(combined_species), big.mark=",")`) of the digitized fauna has DNA barcodes available.  In 2012, it was estimated that, at a global scale, less than 3% of the marine species had been sequenced [@Appeltans2012], and that this number was even lower in some areas (e.g., 2.6% for the Antarctic [@Grant2009]).  This proportion varies by phylum from `r format_output(stat_barcoding$min_eez_value * 100)`% for `r stat_barcoding$min_eez_phylum` to `r format_output(stat_barcoding$max_eez_value * 100)`% for `r stat_barcoding$max_eez_phylum` (Fig. \@ref(fig:proportion-barcoded)). The GOM has comparable coverage in level of DNA barcoding of the fauna to the EEZ, while the PNW is better barcoded with >30% species-level coverage across each of the five most diverse phyla. This is driven in part by the active Canadian barcoding effort in British Columbia.

The above estimates are based on species names, but many traditional species include multiple cryptic species [@Knowlton2000;@Meyer2005-evolution]. For example, up to `r format_output(100 * stat_bold$p_spp_multi_bin)`% of `r format(stat_bold$n_spp_not_singleton, big.mark=",")` species that have DNA barcodes available in the digitized US fauna, are represented by more than 1 record in BOLD include cryptic species, as denoted by assignment to  more than 1 BIN (BOLD Index Number, molecular operational taxonomic units that closely resemble species [@Ratnasingham2013]). Specimens identified under the same species but sampled from distant locations are especially prone to representing cryptic species. While geographic information is missing for most records (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records, big.mark=",")` BOLD records analyzed here have no geographic coordinate information associated with them), of the `r bold_geography_stats$n_spp_with_records` marine invertebrates species from the US with BOLD records, only `r format_output(bold_geography_stats$p_within_eez * 100)`% have at least one record collected within the EEZ and `r format_output(bold_geography_stats$p_in_north_america * 100)`% list a North American country as the site of collection. Although DNA barcodes are available for `r format_output(prop_spp_barcoded$p_barcoded * 100)`% of the marine invertebrate fauna, many of these records have been collected outside the Atlantic and Pacific US biogeographic regions, and therefore geographical differentiation has not been evaluated for most taxa. It is thus crucial to ensure that precise and accurate geographic information is provided when DNA barcodes are published, and that enough samples for each species are sequenced to capture more accurately their geographic range and the extent of their genetic diversity.

@@

Based on the results of this study, we recommend a multi-level approach for both improving the quality of existing data and for the efficient collection of new data. Our recommendations extends the recommendations formulated by Duffy et al. [@Duffy2013], and are informed by the results of our analyses.

@@

This study highlights that we have too little data about too few species of marine invertebrates. We recommend that additional coordinated field expeditions be organized to sample more comprehensively the marine environment. The relationship between sampling effort and diversity from museum collections makes clear that any additional samples will lead to improving our knowledge of species ranges, and will lead to the discovery of many new species. Concerted and collaborative efforts are needed to share expertise and limited resources. Biodiversity surveys that target all animal phyla can be very effective in documenting biodiversity. For instance, one of us (GP) participated in the summer 2017 in a an inventory of Kaneohe Bay, Hawaii, where ~2,400 records representing ~1,300 species were collected in 11 days of field work. While diversity is higher in Hawaii than in the continental US, this is a modest effort that led to the collection of a number of species that represent a few percents of the total diversity of the US EEZ marine invertebrate fauna.

@@

Integrating metabarcoding as part of the data collection conducted in biodiversity surveys is a promising approach to screen an habitat for its potential to lead to new discoveries. For instance, by analyzing metabarcoding samples that contain planktonic larvae from a given habitat, we could get an indication of the taxonomic groups that would benefit most from additional sampling effort, or provide an estimate for the proportion of species that has not been sampled yet. Additionally, data generated by metabarcoding efforts can be re-interpreted as more DNA barcodes associated with correctly identified specimens become available. Building the library of DNA barcodes in parallel with metabarcoding efforts ensures that database matches correspond to actual specimens deposited in natural history collections, ruling out possible artifacts of the methods and permits resolution of specimen identification conflicts.  With such approaches, the geographic distribution of species can build on and be linked to existing records, species and their natural history, rather than being limited to matching DNA sequences across studies.

@@

To generate the combined list of records from iDigBio and OBIS, we removed putative duplicate based on the following criteria: same phylum, same accepted species name from WoRMS, same geographic coordinates (rounded to the 4th decimal), and same date of collection. When the date is missing, a single record per species with identical coordinates is included.

>>>>

---
title: "The state of knowledge for marine invertebrate biodiversity in the continental US"
date: ' ' #'`r format(Sys.time())`'
output:
  bookdown::pdf_document2:
    includes:
    #  before_body: ../template/doc_prefix.tex
      in_header: preamble.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc: no
fontsize: 12pt
link-citations: yes
csl: nature.csl
bibliography: /home/francois/Library/dark-plankton.bib
abstract: Museum collections have been recording our knowledge about species and their geographic ranges for about 250 years, but until recently, this information was fragmented across institutions and difficult to synthesize. The digitization of biodiversity information is now enabling the synthesis of this information, and can be used to highlight knowledge gaps. Here we use publicly accessible biodiversity data to assemble a list of `r format(n_distinct(combined_species$worms_valid_name), big.mark = ",")` species of marine invertebrates from the continental United States. Our knowledge about species richness and their geographic ranges is limited by sampling effort, as `r prop_10_specimens_species`% of the species are known from less than 10 occurrences. While recent developments in high-throughput sequencing promise to greatly speed up the documentation of marine biodiversity, anchoring results from such studies into existing DNA sequence data and associated vouchers would provide a richer biological context to interpret the insights provided by these datasets. However, with `r format_output(prop_spp_barcoded$p_barcoded * 100)`% of the marine invertebrate fauna that has sequences available for the most widely used DNA barcoding gene (cytochrome oxidase I, COI), and given that for many species these records have been sampled in biogeographic areas outside the USA, additional efforts are needed to assemble a comprehensive library of reference DNA barcoding sequences. More sampling, coordinated at the national scale, is needed to improve our knowledge of species and their geographic ranges. This data will allow the detection of the impacts of climate change and other anthropogenic strains on marine biodiversity. Advances in computing, information systems, and sequencing technologies, have set the stage to inform where future efforts for documenting biodiversity should be applied.
---

@@

Several factors may contribute to species being included in the regional lists but not having records in biodiversity databases. The first category of biases is linked to the likelihood of having specimens for these species in digitized museum collections. Rare species may only have been collected a few times, and the corresponding records have not been digitized yet. Species that are difficult to preserve because they are large, very small, or fragile, are deposited in museums less frequently. The other category of biases is associated with data quality for these records. Some of these species may be in the databases but do not have geographic coordinates associated with them, and therefore would not be included in the results of our query. While it is difficult to assess biases associated with the likelihood of preservation of a species, we investigated how missing geographic coordinates are affecting the discrepancies between the checklists and the databases.

@@

The authors of these checklists did not have access to digital databases allowing them to access  museum records across institutions, which explains why some species were not included in the checklists. Many of these records were difficult to discover before digitization which, for the most part, has occurred after the compilation of these checklists. We estimate that using biodiversity databases increases the number of species included in the checklist by `r format_output(spp_not_in_lists$gom_not_in_list/n_distinct(total_gom_species$worms_valid_name)*100)`% for the Gulf of Mexico, and `r format_output(spp_not_in_lists$pnw_not_in_list/n_distinct(total_pnw_species$worms_valid_name)*100)`% for the Pacific Northwest.

@@

The marine invertebrates of the US have generally been characterized genetically at a higher level than other parts of the world: `r format_output(prop_spp_barcoded$p_barcoded * 100)`% (n = `r format(nrow(combined_species), big.mark=",")`) of the digitized fauna has DNA barcodes available.  In 2012, it was estimated that, at a global scale, less than 3% of the marine species had been sequenced [@Appeltans2012], and that this number was even lower in some areas (e.g., 2.6% for the Antarctic [@Grant2009]).  This proportion varies by phylum from `r format_output(stat_barcoding$min_eez_value * 100)`% for `r stat_barcoding$min_eez_phylum` to `r format_output(stat_barcoding$max_eez_value * 100)`% for `r stat_barcoding$max_eez_phylum` (Fig. \@ref(fig:proportion-barcoded)). The GOM has comparable coverage in level of DNA barcoding of the fauna to the EEZ, while the PNW is better barcoded with >30% species-level coverage across each of the five most diverse phyla. This is driven in part by active Canadian barcoding efforts in British Columbia.

The above estimates are based on species names, but many traditional species include multiple cryptic species [@Knowlton2000;@Meyer2005-evolution]. For example, up to `r format_output(100 * stat_bold$p_spp_multi_bin)`% of `r format(stat_bold$n_spp_not_singleton, big.mark=",")` species that have DNA barcodes available in the digitized US fauna, are represented by more than 1 record in BOLD include cryptic species, as denoted by assignment to  more than 1 BIN (BOLD Index Number, molecular operational taxonomic units that closely resemble species [@Ratnasingham2013]). Specimens identified under the same species but sampled from distant locations are especially prone to representing cryptic species. While geographic information is missing for most records (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records, big.mark=",")` BOLD records analyzed here have no geographic coordinate information associated with them), of the `r format(bold_geography_stats$n_spp_with_records, big.mark=",")` marine invertebrates species from the US with BOLD records, only `r format_output(bold_geography_stats$p_within_eez * 100)`% have at least one record collected within the EEZ and `r format_output(bold_geography_stats$p_in_north_america * 100)`% list a North American country as the site of collection. Although DNA barcodes are available for `r format_output(prop_spp_barcoded$p_barcoded * 100)`% of the marine invertebrate fauna, many of these records have been collected outside the Atlantic and Pacific US biogeographic regions, and therefore geographical differentiation has not been evaluated for most taxa. It is thus crucial to ensure that precise and accurate geographic information is provided when DNA barcodes are published, and that enough samples for each species are sequenced to capture more accurately their geographic range and the extent of their genetic diversity.

@@

Based on the results of this study, we recommend a multi-level approach for both improving the quality of existing data and for the efficient collection of new data. Our recommendations extends those formulated by Duffy et al. [@Duffy2013], and are informed by the results of our analyses.

@@

This study highlights that we have too little data about too few species of marine invertebrates. We recommend that additional coordinated field expeditions be organized to sample more comprehensively the marine environment. The relationship between sampling effort and diversity from museum collections makes clear that any additional samples will lead to improving our knowledge of species ranges, and will lead to the discovery of many new species. Concerted and collaborative efforts are needed to share expertise and limited resources. Biodiversity surveys that target all animal phyla can be very effective in documenting biodiversity. For instance, one of us (GP) participated in the summer 2017 in a an inventory of Kaneohe Bay, Hawaii, where ~2,400 records representing ~1,300 species were collected in 11 days of field work. While diversity is higher in Hawaii than in the continental US, this effort led to the collection of a number of species that represent a few percents of the total diversity of the US EEZ marine invertebrate fauna. It demonstrates that documenting the marine fauna of the US is doable.

@@

Integrating metabarcoding as part of the data collection conducted in biodiversity surveys is a promising approach to screen an habitat for its potential to lead new discoveries. For instance, by analyzing metabarcoding samples that contain planktonic larvae from a given habitat, we could get an indication of the taxonomic groups that would benefit most from additional sampling effort, or provide an estimate for the proportion of species that has not been sampled yet. Additionally, data generated by metabarcoding efforts can be re-interpreted as more DNA barcodes associated with correctly identified specimens become available. Building the library of DNA barcodes in parallel with metabarcoding efforts ensures that database matches correspond to actual specimens deposited in natural history collections, ruling out possible artifacts of the methods and permits resolution of specimen identification conflicts.  With such approaches, the geographic distribution of species can build on and be linked to existing records, species and their natural history, rather than being limited to matching DNA sequences across studies.

@@

To generate the combined list of records from iDigBio and OBIS, we removed putative duplicate based on the following criteria: same phylum, same accepted species name from WoRMS, same geographic coordinates (rounded to the 4th decimal), and same date of collection. When the date was missing, a single record per species with identical coordinates is included.
